---
title: "INTRODUCTION TO R: DATA FRAMES"
source: DataCamp
---

### 5.1 What is a data frame?

You will often work with data sets that contain different data types instead of only one. A *data frame* has the variables of a *data set* as columns and the observations as rows.

Load a built-in example data frame in R called `mtcars`:
```{r}
mtcars
```
When you work with large data sets and data frames, it's better to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire data set. `head()` function shows the first observations of a data frame.
```{r}
head(mtcars) 
```
Similarly, `tail()` function shows the last observations of a data frame:
```{r}
tail(mtcars)
```

### 5.2 Structure of a data frame

The structure `str()` function can be used to to get a rapid overview of your data:

-    the total number of observations
-    the total number of variables
-    a full list of the variable names
-    the data type of each variable
-    the first observations

Applying the `str()` function will often be the first thing that you do when receiving a new data set or data frame. It is a great way to get more insight in your data set before diving into the real analysis.
```{r}
str(mtcars) 
```

### 5.3 Create a data frame

Construct a data frame that describes the main characteristics of eight planets in our solar system:
```{r}
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
```
Add the type of planet (Terrestrial or Gas Giant):
```{r}
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
```
Add the planet's diameter relative to the diameter of the Earth:
```{r}
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
```
Add the planet's rotation across the sun relative to that of the Earth:
```{r}
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
```
Add information whether the planet has rings or not (TRUE or FALSE):
```{r}
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)
```
Now create a data frame from those vectors:
```{r}
planets_df <- data.frame(name, type, diameter, rotation, rings)
```
The vectors become different columns of the data frame. Because every column has the same length, the vectors should also have the same length as `planets_df`.

Check the structure of the data frame:
```{r}
str(planets_df)
```

### 5.4 Selection of data frame elements

Similar to vectors and matrices, select elements from a data frame with square brackets `[]`. By using a comma, you can indicate what to select from the rows and the columns respectively.

Print out diameter of Mercury (row 1, column 3):
```{r}
planets_df[1,3] 
```
Print out data for Mars (entire fourth row):
```{r}
planets_df[4,] 
```
Instead of using numerics, it's also possible to use the variable names to select columns of a data frame.
```{r}
planets_df[1:3,2] 
```
Or, it is often easier to just make use of the variable name:
```{r}
planets_df[1:3, "type"]
```
If you want to select all elements of the variable "diameter":
```{r}
planets_df[,3]
```
Or alternatively, use the name column:
```{r}
planets_df[,"diameter"] 
```
But the shorter way is using `$` symbol:
```{r}
planets_df$diameter
```
Select the rings variable from `planets_df`:
```{r}
rings_vector <- planets_df$rings
# See which planet has a ring
rings_vector
```
This means that the first four observations do not have a ring, but the other four do. However, you do not get a nice overview of the names of these planets, their diameter, etc. 

Select all columns for planets with rings by leaving the columns part inside the `[]` empty:
```{r}
planets_df[rings_vector,] 
```
Or create a subset with a condition from the data frame
```{r}
subset(planets_df, subset = rings) 
```

### 5.5 Sorting

In data analysis you can sort data according to a certain variable in the data set. The `order()` function gives the ranked position of each element when it is applied on a variable.
```{r}
a <- c(100, 10, 1000)
order(a)
```
This means we can use the output of `order(a)` to reshuffle `a`:
```{r}
a[order(a)]
```
Rearrange the data frame `planets_df` such that it starts with the smallest planet and ends with the largest one. 
```{r}
# Sorting on "diameter" column
positions <- order(planets_df$diameter)
```
Use positions to sort `planets_df`
```{r}
planets_df[positions,]
```