---
title: "1. INTRODUCTION TO R BASICS"
source: DataCamp
---

#### 1.1 Arithmetic with R

In its most basic form, R can be used as a simple calculator.        

```{r}
# Addition: +
5 + 5 
# Subtraction: -
5 - 5 
# Multiplication: *
3 * 5 
# Division: /
(5 + 5) / 2 
# Exponentiation: ^ (raise the power)
2 ^ 5 
# Modulo: %% (returns the remainder of the division)
28 %% 6 
```

#### 2.2 Variable assignment

A *variable* stores a value or an *object*. You can then later use this variable's name to easily access the value or the object that is stored within this variable.

For example, assign the value 23 to `my_var`:
                 
```{r}
my_var <- 23
```

Print out the value of `my_var`:

```{r}
my_var
```

Suppose you have a fruit basket. Store the number of apples and oranges  in variables, then calcuate the total pieces of fruits you have.

```{r}
my_apples <- 5
my_oranges <- 6
my_apples + my_oranges
```

Assign the result to a new variable `my_fruit`

```{r}
my_fruit <- my_apples + my_oranges
# See how many pieces of fruits in total
my_fruit
```

### 3.3 Basic data types in R

* Decimal values like 4.5 are called *numerics*.
* Natural numbers like 4 are called *integers*. Integers are also numerics.
* Boolean values  (TRUE or FALSE) are called *logical*. 
* Text or string values are called *characters*.

Assign different values to different types of data:

```{r}
my_numeric <- 42
my_character <- "universe"
my_logical <- FALSE
```

Check the data type of a variable with `class()` function:

```{r}
class(my_numeric)
class(my_character)
class(my_logical)
```