---
title: 'INTRODUCTION TO R: FACTORS'
source: DataCamp
---

### 4.1 What is a factor and why we use it?

*Factors* refer to a statistical data type used to store categorical values. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories.

Create a vector that contains all the observations that belong to a limited number of categories:
```{r}
sex_vector <- c("Male", "Female", "Female", "Male", "Male")
```
There're two categories here, or the *factor levels* here are "Male" and "Female".
```{r}
# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)
# Print out to see the factor levels
factor_sex_vector 
```

There are two types of categorical variables: a *nominal* categorical variable and an *ordinal* categorical variable.

A *nominal* variable is a categorical variable without an implied order.
```{r}
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector  
```
It is impossible to say that one is worth more than the other.

In contrast, *ordinal* variables do have a natural ordering. 
```{r}
temperature <-  c("High", "Low", "High", "Low", "Medium")
factor_temperature <- factor(temperature, ordered = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature
```

### 4.2 Factor levels

Some datasets contain factors with specific factor levels. These levels can be changed with `levels()` function for clarity. For example, the sex of the correspondents in a survey were recorded in short as "F" and "M".
```{r}
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
# Specify the levels
levels(factor_survey_vector) <- c("Female", "Male")
# Print out the result
factor_survey_vector
```

### 4.3 Summarize a factor

You would like to know how many "Male" and "Female" responses to your survey. Generate a summary for the survey with `summary()` function:
```{R}
summary(survey_vector) # this shows the length, class and mode of the dataset, not what you expect.
# Generate a summary for the factored survey 
summary(factor_survey_vector) 
```
"Male" and "Female" are nominal (unordered) factor levels and can't be compared.

### 4.4 Ordered factors

Sometimes you will also deal with factors that do have natural ordering. For example, evaluate the performance of 5 data analysts as "slow", "medium" and "fast":
```{r}
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
```
`speed_vector` should be converted into an ordinal factor since its categories have a natural ordering.
```{r}
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))
```

-    By setting the argument `ordered` to `TRUE`, you indicate that the factor is ordered. 
-    With the argument `levels` you give the values of the factor in the correct order.

Now you can compare elements in the vector, for example, compare the performance of data analyst 2 and data analyst 5:
```{r}
da2 <- factor_speed_vector[2]
da5 <- factor_speed_vector[5]
da2 > da5 
```