---
title: 'INTERMEDIATE R: FUNCTIONS'
source: DataCamp
---

### 3.1 Introduction to functions

We already know some buit-in functions in R. For example, create a list with `list()`, or display a variable with `print()` function.

-    Statistical function, like `sd()` for standard deviation:
```{r}
sd(c(1, 5, 6, 7))  # specify function arguments inside () 
# Assign the input vector to a variable, then call sd() function
values <- c(1, 5, 6, 7)
sd(values)         
# Assign the result of the function to a variable to reuse it
my_sd <- sd(values)
```

-    Function documentation:
`help(sd)`
or
`?sd` (look up in the function documentation for more info).

The syntax of `sd()` could be found in the help page
`sd(x, na.rm = FALSE)`  
R matches the values to the `x` argument by position or by name:`sd(x = values)`  

-    `na.rm` argument is a logical value indicating whether or not missing values should be removed. 
```{r}
values <- c(1, 5, 6, 7, NA)
sd(values) 
# set na.rm argument to TRUE
sd(values, TRUE) # or matching by name
sd(values, na.rm = TRUE)
```
`sd()` will return an error if there's no argument is specified.

-    `args()` function: to learn about the arguments of a function without having to go through the entire document.
```{r}
args(sd) # before using an R function, you should clarify which arguments it expects.
```
Consult the documentation on the `mean()` function: `?mean`
```{r}
args(mean)  # inspect the arguments.
```
 
Continue with the exercise on the views on social media:
```{r}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
```
Calculate average number of views:
```{r}
(avg_lin <- mean(linkedin))
(avg_fb <- mean(facebook)) # put the variable inside the parenthese will print it out in the console.
```
Calculate the `mean` of the sum:
```{r}
(avg_sum <- sum(mean(linkedin), mean(facebook)))
```
Using `trim` argument to reduce the decimal ending:
```{r}
(avg_trimmed <- mean(linkedin + facebook, trim = 0.2)) # the use of a trimmed mean helps eliminate the influence of outliers or data points on the tails that may unfairly affect the traditional mean.
```
Suppose there're missing values in the view record:
```{r}
linkedin <- c(16, 9, 13, 5, NA, 17, 14)
facebook <- c(17, NA, 5, 16, 8, 13, 14)
```
To calcuate the average view using `mean()`, set `na.rm()=TRUE`:
```{r}
mean(linkedin, na.rm = TRUE)
# Calculate the mean absolute function
mean(abs(linkedin - facebook), na.rm = TRUE) # Functions inside functions!
```

### 3.2 Writing functions

A *function* typically serves a particular need or solves a particular problem, without having to care about how the function does this. R allows to write your own function and use it as a standard function like `sd()`, `mean()`...  

The syntax for writing a funtion:  
`my_func <- function(arg1, arg2) {` # create a new function  
  `body ` # create a function that takes arguments arg1 and arg2  
`} `      # performs the code in the body on these arguments

Define a function that calculates the triple of its input, called `triple`:
```{r}
triple <- function(x) { # x represents input as a number
  3*x                   # calculates the triple of x
}
```
Check if the new object is already defined in the workspace:
```{r}
ls() # list the object "triple"
```
Calculate the triple of 6:
```{r}
triple(6) # numeric 6 matched to argument(x)
          # function body is executed: 3 * 6
          # last expression evaluated becomes the return value.
```
-    `return()` statement: explicitly specify the return value.
```{r}
triple <- function(x) {
  y <- 3*x
  return(y)
}         # returns the same result.
```
Using a `return` statement at the end of the function is not always useful, but there are other cases where it becomes handy.

R passes arguments by value.  
```{r}
a <- 5
triple(a)
a
```
The R objects you pass to a function can never change unless you do an explicit assignment.
```{r}
increment <- function(x, inc = 1) {
  x <- x + inc
  x
}
count <- 5
a <- increment(count, 2)
b <- increment(count)
count <- increment(count, 2)
```
R passes arguments by *value* and not by *reference*, the value of count is not changed after the first two calls of `increment()`. Only in the final expression, where count is re-assigned explicitly, does the value of `count` change. 

Write a function called `math_magic()` that takes 2 numbers as input and calculates the sum of the product and the division of both numbers:
```{r}
math_magic <- function(a, b) {
  a*b + a/b
}
```
Sourcing the new function with numbers 4 and 2:
```{r}
math_magic(4,2) 
```
R will return an error if the input has only one argument:
`math_magic(4)` 

Making the second argument optional:
```{r}
math_magic <- function(a, b = 1) { # adding default value
  a*b + a/b
}
```
Sourcing the function again:
```{r}
math_magic(4) # because argument b is not specified, R set it to default value that equals 1.
```
Call `math_magic()` with numbers 4 and 0
```{r}
math_magic(4,0) # result is Inf, or infinity, because 4/0 = inf
```
Extend the function with an *if-return* statement to guard the function agains this misuse:
```{r}
math_magic <- function(a, b = 1){
  if(b == 0){
    return(0)
  }
  a*b + a/b
}
```
Recall the function again:
```{r}
math_magic(4,0) # return statement can halt the execution anytime.
```
Create a function that takes 2 arguments and returns the sum of the absolute values of both arguments:
```{r}
sum_abs <- function(a, b){
  abs(a) + abs(b)
}
sum_abs(-2, 3)
```
There are situations in which the function does not require an input. For example, write a function that gives the random outcome of throwing a fair die:
```{r}
throw_die <- function(){
  number <- sample(1:6, size = 1)
  number
}
throw_die()
```
Define a function without argument that prints out "Hi there!" and returns `TRUE`:
```{r}
hello <- function(){
  print("Hi there!")
  return(TRUE)
}
hello()
```

Create a function that takes one argument and returns that number squared:
```{r}
power_two <- function(x){
  x^2
}
```
Add an optional argument named `print_info`, that is `TRUE` by default and and `if` statement.
```{r}
power_two <- function(x, print_info = TRUE){
  y <- x^2
  if(print_info){
    print(paste(x, "to the power two equals", y))
  }
  return(y)
}
```
Call the newly defined function with different inputs:
```{r}
power_two(5) 
power_two(5, FALSE)
power_two(5, TRUE)
```
-    Function scoping: variables that are defined inside a function are not accessible outside that function.
```{r}
two_dice <- function(){
  possibilities <- 1:6
  dice1 <- sample(possibilities, size = 1)
  dice2 <- sample(possibilities, size = 1)
  dice1 + dice2
} # whatever the way of calling the two_dice() function, R won't have the access to dice1 and dice2 outside of the function.
```

Continue the exercise with the record of social media views in `linkedin` and `facebook`. Write a function that interprets the view of a singe record day:
```{r}
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}
```
Call the interpret function twice:
```{r}
interpret(linkedin[1])
interpret(linkedin[2])
```
Now write a function that interprets all the data from the daily views.
Make `return_sum` an optional argument, that is `TRUE` by default.
```{r}
interpret_all <- function(views, return_sum = TRUE) {
  count <- 0
  for (v in views) {
count <- interpret(v) + count
  }
  if (return_sum) {
return(count)
  } else {
return(NULL)
  }
}
```
Call the `interpret_all()` function on both `linkedin` and `facebook`.
```{r}
interpret_all(linkedin)
```
### 3.3 Anonymous functions 

Functions in R are *objects* in their own right. This means that they aren't automatically bound to a name. When you create a function, you can use the assignment *operator* to give the function a name. It's perfectly possible, however, to not give the function a name. This is called an *anonymous function*.

-    Named function:
```{r}
triple <- function(x) { 3 * x }
```
- Anonymous function with same implementation
```{r}
function(x) { 3 * x }
```
Anonymous functions are helpful when you need to use the function only once, avoid defining it with a name. 