---
title: "INTERMEDIATE R: PACKAGES"
source: DataCamp
---

R packages are bundles of code, data, documentation, and tests that are easy to share with others. For example, the `mean()`, `list()` and `sample()` functions are part of the **Base R** package, which contains the basic functionality to use R.

### 4.1 Install packages
Base R package is automatically installed and loaded. The package for data visualization, for example, has to be installed and loaded manually: `install.packages("ggvis")`. This goes to CRAN, download and install the package in your system.

### 4.2 Load packages
When R loads a package, it actually attaches it to the search list.
```{r}
search() # list of packages and environments that R looks through to find the variable or function you want to use.
```
*GlobalEnv* is the workspace where the user-defined R objects are stored.  
Try to access a function in the installed **ggvis** packages:
`ggvis(mtcars, ~wt, ~hp)` returns an error, because **ggvis** is not yet in the search list.   
Load ggvis manually:
```{r}
library("ggvis")
```
Check once more:
```{r}
search() # ggvis is now a part of the list.
```
Run the ggvis function again:
```{r}
ggvis(mtcars, ~wt, ~hp) # print out a graph.
```

Another way to load packages is to use `require()` function for loading packages that are not yet installed.
```{r}
require("data.table") # return a warning if attaching package fails.
```

Check if the package is already attached:
```{r}
(result <- require("data.table")) 
```
This is a good alternative to avoid errors when attaching dynamically packages inside functions.

Load ggplot2 package:
```{r}
require("ggplot2")
```
Build a plot of two variables of the `mtcars` data frame using quick plot function from the package **ggplot2**
```{r}
qplot(mtcars$wt, mtcars$hp)
```