---
title: "INTERMEDIATE R: THE APPLY FAMILY"
source: DataCamp
---

### 5.1 The lapply
When using R, you will encounter *vectors* and *lists* containing all sorts of information. We could use `for` *loop* to iterate over all kinds of data structures. But there's even an easier way with `lapply`.  
Reconstruct again the previous list example:
```{r}
nyc <- list(pop = 8405837, 
            boroughs = c("Manhatten", "Bronx", "Brooklyn", "Queens",                          "Staten Islands"), 
            capital = FALSE)
```
Suppose you want to find out the class of each element in this list.
```{r}
for(info in nyc) {
  print(class(info))
}
```
The `for` loop is, however, a long code just to find out the class of the elements in the list. Instead we could use `lappy()`.
```{r}
lapply(nyc, class)
```
Another example with city names:
```{r}
cities <- c("New York", "Paris", "London", "Tokyo", "Rio de Janiero", "Cape Town")
```
Suppose we want to build a vector of the same length, containing the number of characters of each city name.
```{r}
num_chars <- c() # create an empty vector
for(i in 1:length(cities)) { # using loop index
  num_chars[i] <- nchar(cities[i]) # use nchar() function.
}
num_chars # we got the result, but it's a lot of code.
```
Refactoring this code using `lapply()` is easy.
```{r}
lapply(cities, nchar) # result a list containing the number of characters of the corresponding cities in the input vector.
```
`lapply` always returns a list, irrespective of the input data structure. By using `lapply`, the code looks much more intuitive and readable.  
`lapply` can also be used not just with default functions but customized functions.  
Assume we have a list of oil prices per gallon:
```{r}
oil_prices <- list(2.37, 2.49, 2.18, 2.22, 2.47, 2.32)
```
Write a function that triples every element of `oil_prices`:
```{r}
triple <- function(x) {
  3 * x
}
```
Apply this `triple` function on each element using `lapply()`:
```{r}
lapply(oil_prices, triple)
```
Try to make this triple function more generic:
```{r}
multiply <- function(x, factor) { # use an additional argument.
  x * factor
}  # we can now choose which factor to multiply the input.
```
Use `multiply` inside the `lapply()` function:
```{r}
lapply(oil_prices, multiply, factor = 4)
```
To put it generally, `lapply` takes a vector or a list `x`, and applies a specified function to its members. If the function requires additional arguments, it could be passed inside the `lapply()` function.  
**Syntax**: `lapply(x, FUNC, ...)`   
The output of `lapply()` is a list, where each element is the result of appply `FUNC` on the corresponding element of `x`. 

Create a vector or famous mathematicians/statisticians and the year they were born:
```{r}
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
```
Split names from birth year using `strplit()` fucntion:
```{r}
split_math <- strsplit(pioneers, split = ":") # split the strings on the : sign.
```
Convert to lowercase strings using `tolower()` function:
```{r}
split_low <- lapply(split_math, tolower) 
```
Inspect the contents of `split_low`:
```{r}
str(split_low) # names are in converted to lowercase now.
```
Now write new functions that allows to select the names and the years separately:
```{r}
select_first <- function(x){
  x[1]
}
select_second <- function(x){
  x[2]
}
```
Apply the new functions:
```{r}
names <- lapply(split_low, select_first)
years <- lapply(split_low, select_second)
```
Or generic select function:
```{r}
select_el <- function(x, index) {
  x[index]
}
```
Use `lapply()` with additional arguments:
```{r}
names <- lapply(split_low, select_el, index = 1)
years <- lapply(split_low, select_el, index = 2)
```

### 5.2 The sapply
Using the `cities` example again, `lapply()` returns a list, although the result could fit nicely into a vector. There's an easier way to tackle the case in which all the results have the same type by using `sapply()`.
```{r}
sapply(cities, nchar) # short for "simplified apply".
```
Under the hood, something more complex is going on:  

-    `sapply` calls `lappy` to apply the `nchar` function over each element of the cities vector.  
-    Then it uses the `simplify2array` function to convert that list `lapply` generated to an *array*.  
-    The `sapply` in this case converts the result to a one dimensional array, which is a *vector*. 

`sapply` even found a sensible way of naming this vector. This argument could be turned off.
```{r}
sapply(cities, nchar, USE.NAMES = FALSE) # TRUE by default.
```
What happens if the function you want to apply over the input, each time returns a vector containing two values instead of one?
```{r}
first_and_last <- function(name) {
  name <- gsub(" ", "", name)
  letters <- strsplit(name, split = "")[[1]]
  c(first = min(letters), last = max(letters))
} 
```
The written function splits up a string to its letters, and returns the minimum and maximum letter according to the alphabetial order.
```{r}
first_and_last("New York")
```
Use this function for every city name in `cities`:
```{r}
sapply(cities, first_and_last) # the result is now a matrix.
```
What if this simplification is not possible?  
There are cases in which the function does not always return a vector of the same length at all times. For this case, simplification to a vector.

Write a function that returns a vector of all the letters that are used inside a character string:
```{r}
unique_letters <- function(name) {
  name <- gsub(" ", " ", name)
  letters <- strsplit(name, split = "")[[1]]
  unique(letters)
} 
unique_letters("London") #  returns a vector containing the unique letters in "London".
```
First see how  the `cities` behave on your function:
```{r}
lapply(cities, unique_letters) # returns a list of vectors with different length.
```
Trying to simplify this list could lead to strange result.
```{r}
sapply(cities, unique_letters) # also returns a list because R couldn't think of a meaningful way of symplifying the result.
```
Create a list that contains the temperature measurements for 7 days:
```{r}
temp <- list(c(3, 7, 9, 6, -1), c(6, 9, 12, 13, 5), c(4, 8, 3, -1, -3), c(1, 4, 7, 2, -2), c(5, 7, 9, 4, 2), c(-3, 5, 8, 9, 4), c(3, 6, 9,  4, 1)) # temp is a list of length 7, while each element is a vector of length 5.
```
Use `lapply()` to find each day's minimum temperature:
```{r}
lapply(temp, min) # return a list of length 7
```
Use `sapply()` instead:
```{r}
sapply(temp, min) # returns a vector of length 7.
```
Similary, find out each day's maximum temperature:
```{r}
sapply(temp, max)
```
Create a function to calculate the average of the minimum and the maximum temperatures of the vector:
```{r}
extremes_avg <- function(x) {
  (min(x) + max(x)) / 2
}
```
Apply `extremes_avg()` over `temp` using `sapply()`:
```{r}
sapply(temp, extremes_avg)
```
Create a function that returns min and max of a vector:
```{r}
extremes <- function(x) {
  c(min = min(x), max = max(x))
}
```
Apply `extreme()` over `temp` using `sapply()`:
```{r}
sapply(temp, extremes)
```
In some cases, `sapply()` can't simplify the output and will return a list like `lapply()`.
```{r}
below_zero <- function(x){
  return(x[x < 0])
}
freezing_s <- sapply(temp, below_zero)
freezing_l <- lapply(temp, below_zero)
```
Use `identical()` function to compare `freezing_s` and `freezing_l`:
```{r}
identical(freezing_s, freezing_l) 
```
-    `sapply` with functions that return `NULL`  
Create a function `print_info()` that takes a vector and prints the average of this vector:
```{r}
print_info <- function(x) {
  cat("The average temperature is", mean(x), "\n")
}
```
`cat()` is simlar like `print()` but less conversion, and `cat()` returns invisibly `NULL`.
```{r}
sapply(temp, print_info)
```

### 5.3 The vapply
`vapply` is similar to `sapply`. It also uses the `lapply` and then tries to simplify the result. However, when using `vapply`, we have to explicitly say what the type of the return value will be. In `sapply`, this is not required nor possible.  
**Syntax**: `vapply(x, FUN, FUN.VALUE,..., USE.NAME = TRUE)`  
The syntax is similar to `sapply`'s, but the `FUN.VALUE` is new. This argument should be a general template for the return value of `FUN`, the function that you want to apply over the input `x`. 
Call the `vapply` on the `cities` example:
```{r}
vapply(cities, nchar, numeric(1)) # numeric(1) tells the vapply() that nchar() should return a single numerical value.
```
The result is exactly the same as using `sapply()`. However, the pre-specification of `FUN`'s returns value makes vapply a safer alternative to sapply.  
Reuse the written function `first_and_last()` with `vapply`:
```{r}
vapply(cities, first_and_last, character(2)) # return the same result as sapply.
```
`vapply(cities, first_and_last, character(1))` results an error, as the output of first_and_last() is not expected.
`vapply(cities, first_and_last, numeric(2))` also results an error.

The little extra work in defining the `FUN.VALUES` arguments has the benefits that you really have to think about what your function will return without blindly assuming that the `sapply()` with handle every case for you.  
Use the `temp` example again with a new defined function:
```{r}
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}
vapply(temp, basics, numeric(3))
vapply(temp, function(x, y) {mean(x) >y}, y = 5, logical(1))
```