---
title: "INTERMEDIATE R: DATES AND TIME"
source: DataCamp
---

In R, dates are represented by `Date` objects, while times are represented by `POSIXct` objects. Under the hood, however, these dates and times are simple numerical values.

### 7.1 Dates
Simply ask R what the current date is:
```{r}
today <- Sys.Date()
today
```
See what type of data *today* available:
```{r}
class(today)
```
`Date`: a special R object that represents dates.  
See what `today` looks like under the hood:
```{r}
unclass(today)
```
Create `Date` objects:
```{r}
my_date <- as.Date("1971-05-14") # ISO date format by default.
```
Another date format has to be defined.
```{r}
my_date <- as.Date("1971-14-05", format = "%Y-%d-%m")
my_date
```
In addition to creating dates, you can also convert dates to character strings that use a different date notation. For this, you use the `format()` function. 

-    `%Y`: 4-digit year (1971)  
-    `%y`: 2-digit year (71)  
-    `%m`: 2-digit month (05)
-    `%d`: 2-digit day of the month (14)  
-    `%A`: weekday (Wednesday)  
-    `%a`: abbreviated weekday (Wed)  
-    `%B`: month (January)  
-    `%b`: abbreviated month (Jan)  
```{r}
format(Sys.Date(), format = "%d %B, %Y")
format(Sys.Date(), format = "Today is a %A!")
```
Definition of character strings representing dates:
```{r}
str1 <- "May 23, '96"
str2 <- "2012-03-15"
str3 <- "30/January/2006"
```
Convert the strings to dates: 
```{r}
date1 <- as.Date(str1, format = "%b %d, '%y")
date2 <- as.Date(str2)
date3 <- as.Date(str3, format = "%d/%B/%Y")
```
Convert dates to formatted strings:
```{r}
format(date1, "%A")
format(date2, "%d")
format(date3, "%b %Y")
```
*Date arithmetic*: R performs the calculations using the underlying numerical values, and then converts the result back to human-readable time information again.
```{r}
my_date + 1 # 1 day increment.
```
If we want to know the difference between 2 dates:
```{r}
my_date2 <- as.Date("1998-09-29")
my_date2 - my_date
```

For example, to control your eating habits, you decided to write down the dates of the last five days that you ate pizza.
```{r}
day1 <- as.Date("2020-07-29")
day2 <- as.Date("2020-07-31")
day3 <- as.Date("2020-08-05")
day4 <- as.Date("2020-08-11")
day5 <- as.Date("2020-08-16")
pizza <- c(day1, day2, day3, day4, day5)
```
Difference between last and first pizza day
```{r}
day5 - day1
```
Use the function `diff()` on pizza to calculate the differences between consecutive pizza days.
```{r}
day_diff <- diff(pizza)
```
Average period between two consecutive pizza days:
```{r}
mean(day_diff)
```

Another example, create a vector which contains character strings representing the dates on which the 4 astronomical seasons start:
```{r}
astro <- c(spring = "20-Mar-2015", summer = "25-Jun-2015", fall = "23-Sep-2015", winter = "22-Dec-2015" )
```
Create another vector with the meteorological beginnings of a season:
```{r}
meteo <- c(spring = "March 1, 15", summer = "June 1, 15", fall = "September 1, 15", winter = "December 1, 15")
```
Convert to `Date` object:
```{r}
astro_dates <- as.Date(astro, format = "%d-%b-%Y")
meteo_dates <- as.Date(meteo, format = "%B %d, %y")
```
Calculate the maximum absolute difference between `astro_dates` and `meteo_dates`:
```{r}
max(abs(astro_dates - meteo_dates))
```

### 7.2 Time
Get current time in R:
```{r}
now <- Sys.time()
now
class(now)
```
`POSIXct`: apart form providing rich functionality for calculus and formatting, this class makes sure that the dates and time in R are compatible across different operating systems, according to the *POSIX* standard.
See what now looks like under the hood:
```{r}
unclass(now)
```
Create a `POSIXct` object:
```{r}
my_time <- as.POSIXct("1971-05-24 11:25:15")
```
Similar to working with dates, `as.POSIXct()` is used to convert from a character string to a `POSIXct` object, and `format()` to convert from a POSIXct object to a character string.

-    `%H`: hours as a decimal number (00-23)  
-    `%I`: hours as a decimal number (01-12)  
-    `%M`: minutes as a decimal number  
-    `%S`: seconds as a decimal number  
-    `%T`: shorthand notation for the typical format %H:%M:%S  
-    `%p`: AM/PM indicator 

For a full list of conversion symbols, ask `?strptime` in Help.

Definition of character strings representing times:
```{r}
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"
```
Convert the strings to `POSIXct` objects:
```{r}
time1 <- as.POSIXct(str1, format = "%B %d, %y hours:%H minutes:%M seconds:%S")
time2 <- as.POSIXct(str2) 
```
Convert times to formatted strings:
```{r}
format(time1, "%M")
format(time2, "%I:%M %p")
```
`POSIXct` arithmetic:computaions with `POSIXct` happens in exact the same fashion.
```{r}
my_time + 1 # 1 second increment.
# Calculate the difference between times:
my_time2 <- as.POSIXct("1974-07-14 21:11:55 CET")
my_time2 - my_time # because the time difference is so large, R simply displays the time difference in days.
```
Calculations using `POSIXct` objects are completely analogous to those using `Date` objects.
```{r}
now + 3600          # add an hour
now - 3600 * 24     # subtract a day
```
Adding or substracting time objects is also straightforward.
```{r}
birth <- as.POSIXct("1879-03-14 14:37:23")
death <- as.POSIXct("1955-04-18 03:47:12")
einstein <- death - birth
einstein
```
You're developing a website that requires users to log in and out. You want to know what is the total and average amount of time a particular user spends on your website. This user has logged in 5 times and logged out 5 times as well. 
```{r}
login <- as.POSIXct(c("2020-08-02 10:18:04 UTC", "2020-08-07 09:14:18 UTC", "2020-08-07 12:21:51 UTC", "2020-08-07 12:37:24 UTC", "2020-08-09 21:37:55 UTC"))
logout <- as.POSIXct(c("2020-08-02 10:56:29 UTC", "2020-08-07 09:14:52 UTC", "2020-08-07 12:35:48 UTC", "2020-08-07 13:17:22 UTC", "2020-08-09 22:08:47 UTC"))
```
Calculate the difference between `login` and `logout`:
```{r}
time_online <- logout - login
```
Calculate the total time online:
```{r}
sum(time_online)
```
Calculate the average time online:
```{r}
mean(time_online)
```