---
title: 'INTERMEDIATE R: REVIEW'
source: DataCamp
---

## **1. CONDITIONALS & CONTROL FLOW**

### 1.1 Grade analysis in R
Suppose you're a student in a class with other 199 classmates. You want to analyze the grades of this academy. The highest possible grade is 100, the lowest possible grade is 0.   
You know your own grade.

```{r}
me <- 89
```
You have the access to the anonymized grades of other students.
`other_199 <- scan("other_199.txt")`  
Set directory path to `0_data` and import the text file as a vector with `scan()` function.  
You also have the access to the grades of previous 4 years. Import this data and convert it into a matrix with 4 columns: 
`previous_4 <- matrix(scan("previous_4.txt"), ncol = 4)`  
To format the data so that it'd be easier to work with and analyze later on, merge the three datasets:
```{r}
my_class <- c(me, other_199)
```
Build a 200 x 5 matrix using `cbind()` function to merge columns:
```{r}
last_5 <- cbind(my_class, previous_4)
```
Name last_5 approprieately from `year_1` to `year_5`:
```{r}
colnames(last_5) <- paste0("year_", 1:5)
```
`paste0()` function uses an empty string to join strings.

### 1.2 Explore your data
To get a good feel for your data, it's a good idea to make some visualizations and make some summaries.  
Build a histogram to have an overview of the grade this year:
```{r}
hist(my_class) # your grade (89) is in the 85-90 bin.
```

Distinguish different bars with different colors:
```{r}
hist(my_class, col = rainbow(10)) # use rainbow palette.
```

Generate a summary of each year:
```{r}
summary(last_5)
```
A boxplot is more informative here.
```{r}
boxplot(last_5) # the plot shows only a slight differenct among 5 years.
```

### 1.3 Basic queries
Using *relational operators* to write queries to answer question about `me`, `my_class` and `last_5`.    

-    Is your grade equal to 72?
```{r}
me == 72
```
-    Which grades in your class are higher than 75?
```{r}
my_class > 75
```
-    Which grades in the last 5 years are below or equal to 64?
```{r}
last_5 <= 64
```
The result of the queries should be a vector or a matrix of logicals.

### 1.4 Logical operator
Next to relational operators, there are also *logical operators*, to combine logicals: `&`(AND), `|` (OR), `!` (NOT). 

-    Is your grade greater than 87 and smaller than or equal to 89?
```{r}
me > 87 & me <= 89
```
-    Which grades in your class are below 60 or above 90?
```{r}
my_class < 60 | my_class >90
```

### 1.5 Build aggregates
Answering the question "which grades in your class are higher than 75?" with a vector of logicals is not very insightful. It's much better to ask the question "how many grades in your class are higher than 75?" instead.  
For a logical vector, each `TRUE` counts as **1**, and each `FALSE` as **0**. So summing all the elements of a logical vector is equivalent to counting all the **1**’s (i.e. counting all the `TRUE` values).

-    How many grades in your class are higher than 75?
```{r}
sum(my_class > 75)
```
-    How many students in your class scored strictly higher than you?
```{r}
sum(my_class > me)
```
Just in the same way, use `mean()` to find proportion of `TRUE` values in a logical vector: `mean()` returns the number of `TRUE` values divided by the total number of values.  
-    What's the proportion of grades below or equal to 64 in the last 5 years?
```{r}
mean(last_5 <= 64)
```
-    What's the proportion of grades in your class that are greater than or equal to 70 and lower than or equal to 85?
```{r}
mean(my_class >= 70 & my_class <= 85)
```
-    How many students in the last 5 years had a grade of 80 or 90?
```{r}
sum(last_5 == 80 | last_5 == 90)
```

### 1.6 if, else
Recall the syntax of **if-else**:
```{r}
if (me > 80) {
  print("Good student!")
} else {
  print("Better luck nex year!")  
}
```
Your grade `me` = 89, so the condition in the `if` statement evaluates to `TRUE`, and `print("Good student!)` is executed.
Assign to `n_smart` the number of grades in `my_class` that are greater than or equal to 80.
```{r}
n_smart <- sum(my_class >= 80)
# Write an if-else statement:
if (n_smart > 50) {
  print("smart class")
} else {
  print("rather average")
}
```

### 1.7 else if
Assign `prop_less` the proportion of students whose grade - stored in `my_class` - was lower than yours:
```{r}
prop_less <- mean(my_class < me)
```
Customize further the **if-else** constructs with `else if` statement: 
```{r}
if (prop_less > 0.9) {
  print("you're among the best 10 percent")
} else if (prop_less > 0.8) {
  print("you're among the best 20 percent")
} else {
  print("need more analysis")
}
```
Imbedded **if-else** clauses:
```{r}
if (mean(my_class) < 75) {
  if (mean(my_class) > me) {
    print("average year, but still smarter than me")
  } else {
    print("average year, but I'm not that bad")
  } 
} else {
  if (mean(my_class) > me) {
    print("smart year, even smarter than me")
  } else {
    print("smart year, but I am smarter")
  }
}
```
See whether there were more high achievers than low achievers in your class:
```{r}
top_grades <- my_class[my_class >= 85]
worst_grades <- my_class[my_class < 65]
if (sum(top_grades) > sum(worst_grades)){
  print("top grades prevail")
}