---
title: "IMPORTING EXCEL DATA"
source: DataCamp
---
Excel is a widely used data analysis tool. There are already a lot of packages interacting with Excel to deal with the data in R. 

### 2.1 readxl
For most data related tasks, Excel file contains different sheets that contain tabular data. For example, this *cities.xlsx* contains two sheets containing data of total population of some capitals for two different years.

<img src = "https://raw.githubusercontent.com/Lsinfos/R-00_Base-R/master/2_resources/import_excel.JPG" width = "300">

When you work in R, you typically want to explore your excel file first and then import some data from it.
The **readxl** package developed by Hadley Wickham makes it easy to get data out of Excel and into R.
```{r}
library(readxl)
```
**readxl** basically contains two main functions: 

-   `excel_sheet`: lists the different sheets in the Excel file.
-   `read_excel`: imports the sheet data into R.

**readxl** can handle both *.xls* as *.xlsx* files.
Before you can start importing from Excel, you should find out which sheets are available in the workbook. 
```{r}
excel_sheets("urbanpop.xlsx")
```
This dataset contains urban population metrics for practically all countries in the world throughout time (Source: [Gapminder](https://www.gapminder.org/)). It contains three sheets for three different time periods. In each sheet, the first row contains the column names.
We already know the sheet names now, but that's just the names, not the actual population data. `read_excel` imports the sheet data into R. By default the first sheet in the Excel file is imported as a tibble.
```{r}
read_excel("urbanpop.xlsx")
```
Explicitly tell read_excel() which sheet to import: 
```{r}
read_excel("urbanpop.xlsx", sheet = 3)
```
The `sheet` argument works both with index or the sheet name.
```{r}
read_excel("urbanpop.xlsx", sheet = "1975-2011")
```
However, loading in every sheet manually and then merging them in a list can be quite tedious. Automate this with `lapply()`:
```{r}
pop_list <- lapply(excel_sheets("urbanpop.xlsx"), 
                   read_excel, 
                   path = "urbanpop.xlsx")
```
Display the structure of `pop_list`:
```{r}
str(pop_list)
```
Next to the `path` and `sheet` arguments, there are also other arguments that you can specify in `read_excel()`.

-    `col_names`: by default is TRUE, denoting whether the first row in the Excel sheets contains the column names. If this is not the case, you can set `col_names` to `FALSE`. In this case, R will choose column names for you. You can also choose to set `col_names` to a character vector with names for each column. 
Import the *urbanpop_nonames.xlsx* file which contains the same data as *urbanpop.xlsx* but has no column names in the first row of the excel sheets:
```{r}
read_excel("urbanpop_nonames.xlsx", 
           sheet = 1, 
           col_names = FALSE)
```
Import the file again using a character vector for column names:
```{r}
cols <- c("country", paste0("year_", 1960:1966))
read_excel("urbanpop_nonames.xlsx", 
           sheet = 1, 
           col_names = cols)
```
Excel can contain different data types, such as text, numerics and dates. How this converts to R classes is controlled in the `col_types` argument. By default it is `NULL`, which has R guess the data types of the different columns. But you can also manually specify the `col_types`.
```{r}
read_excel("urbanpop_nonames.xlsx", 
           sheet = 1, 
           col_names = cols,
           col_types = rep("text", 8)) # import all columns as text.
```
Other keywords to enforce a type are "numerics", "date", and "blank". If you use "blank", you're simply ignoring that column.
```{r}
read_excel("urbanpop_nonames.xlsx", 
           sheet = 1, 
           col_names = cols,
           col_types = c("text", rep("blank", 7))) 
```
The rest of the columns are missing. This "blank" keyword can come in handy when you have an Excel file with a lot of columns and you only need some of them in R. 

`skip` argument specifies the number of rows in the Excel sheets R has to skip before actually importing the data.
```{r}
read_excel("urbanpop.xlsx", 
          sheet = 2, 
          col_names = FALSE, 
          skip = 21)
```
`n_max` specifies the number of records to read. 
```{R}
read_excel("urbanpop.xlsx", 
          sheet = 2, 
          col_names = FALSE, 
          n_max = 3)
```
`excel_sheets()` and `read_excel()` are the two functions and the customization possibilities are all you need to get started with Excel data in R. On top of that, **readxl** is extremely fast.

### 2.2 gdata
One alternative to **readxl** is **gdata**, currently mainatained by Gregory Warnes. **gdata** package is an entire suite of tools for performing data manipulation in all sorts of fields. It supercharges the basic R distribution to make handling the data less painful.
```{r}
library(gdata)
```
Among these tools the function `read.xls()` allows you to import Excel data. **gdata** can only handle *.xls* format, but it is easy to install a driver to support the newer *.xlsl* format, as well.
```{r}
urban_pop <- read.xls("urbanpop.xls", sheet = 2)
head(urban_pop)
```
There is no function to list the different sheets in an excel file, like `excel_sheets` in **readxl**.
To import data from Excel, **gdata** uses *Perl*, another interpreted programming language. The Perl code converts the data in the excel sheet to a csv file, and then reading that csv file with the `read.csv()` function that is loaded by default in R, through the **utils** package. This means that all the options that you can specify in `read.csv()`, can also be specified in `read.xls()`. 

<img src = "https://raw.githubusercontent.com/Lsinfos/R-00_Base-R/master/2_resources/import_gdata.JPG" width = "300">

In this respect, the `read.xls()` is an extension of the data import function from the **utils** package to Excel files. This makes it easy to use for people who are familiar with the import functions from the **utils** package and their arguments. On the other hand, the two steps of converting is extremely inefficient. Especially when you have to deal with a huge Excel file; first you have to convert the entire file, then read it.  
```{r}
columns <- c("country", paste0("year_", 1967:1974))
urban_pop <- read.xls("urbanpop.xls", sheet = 2,
         skip = 50, header = FALSE, stringsAsFactors = TRUE,
         col.names = columns)
head(urban_pop)
```
After reading Excel data, we can try to clean and merge it. 
```{r}
path <- "urbanpop.xls"
urban_sheet1 <- read.xls(path, sheet = 1)
urban_sheet2 <- read.xls(path, sheet = 2)
urban_sheet3 <- read.xls(path, sheet = 3)
```
Combine sheets with `cbind()`. Make sure the first column of `urban_sheet2` and `urban_sheet3` are removed to avoid duplicating columns.
```{r}
urban <- cbind(urban_sheet1, 
               urban_sheet2[-1], 
               urban_sheet3[-1])
```
Remove all rows with `NA`:
```{r}
urban_clean <- na.omit(urban)
```
Print out a summary of `urban_clean`:
```{r}
summary(urban_clean)
```

### 2.3 XLConnect
**XLConnect** is one of the most comprehensive packages for working with Excel files through R. written and maintained by Martin Studer. The package is written and maintained by Martin Studer. 
**XLConnect** works as a bridge between Excel and R. You can do practically any action you could do within Excel inside R. **XLConnect** can work with both *xls* and *xlsx* files, and has easy-to-use functions. To get this all working so smoothly, **XLConnect** depends on *Java*. By installing **XLConnect**, it also installs the dependencies **XLConnectJars** package, which contains Java files and class definition that **XLConnect** depends on it. The **rJava** package will also be installed, providing a low-level R to Java interface that XLConnect uses.
```{r}
library(XLConnect)
```
First step is loading a workbook into R with `loadWorkbook()`.
```{r}
book <- loadWorkbook("cities.xlsx")
str(book)
```
This results in a so-called *workbook* object, which is the actual "bridge" between R and Excel. After bulding a workbook object in R, you can use it to get information on the Excel it links to.
Use `getSheets()` to get the names of the sheets:
```{R}
getSheets(book)
```
This works exactly like the `excel_sheets` function from the **readxl** package. The result is a character vector containing the two sheet names.
Apart from sheet information, you can also read the actual data from the sheets, like `read_excel()` from **readxl** and `read.xls()` from **gdata**.
```{r}
readWorksheet(book, sheet = "year_2000")
```
With `readWorksheet()` you can easily specify from which row and which column to start reading information. For example, you want to get the population information for Berlin and Madrid.

<img src = "https://raw.githubusercontent.com/Lsinfos/R-00_Base-R/master/2_resources/import_XLConnect.JPG" width = "200">

```{R}
readWorksheet(book, sheet = "year_2000",
              startRow = 3,
              endRow = 4,
              startCol = 2,
              header = FALSE) # because the first row is skipped.
```
This package is an easy tool to modify the content of your workbook comfortably through R. For example, if want to add fresh population data from 2010, that is stored in a data frame, `pop_2010`.
```{r}
pop_2010 <- data.frame(Capital = c("New York", "Berlin", "Madrid", "Stockholm"), Population = c(8191900, 3460725, 3273000, 1372565))
pop_2010
```
To create a sheet with the data of `pop_2010` into the current workbook, first we create a new empty sheet:

<img src = "https://raw.githubusercontent.com/Lsinfos/R-00_Base-R/master/2_resources/import_XLConnect%202.JPG" width = "300">

```{r}
createSheet(book, name = "year_2010")
```
Next we write the data from `pop_2010` to this new sheet by specifying the name of the new sheet or the number of the sheet:
```{r}
writeWorksheet(book, pop_2010, sheet = "year_2010") # or sheet = 3
```
Finally you have to explicitly save the entire workbook to a file for the changes to take effect. It is recommended to save it with a new file to avoid overwriting the original file.
```{r}
saveWorkbook(book, file = "cities2.xlsx")
```
The `renameSheet()` allows you to rename all the sheets. The first argument is the workbook, then pass the old name and the new name:
```{r}
renameSheet(book, "year_1990", "Y1990")
renameSheet(book, "year_2000", "Y2000")
renameSheet(book, "year_2010", "Y2010")
```
Save the result again to a new file:
```{r}
saveWorkbook(book, file = "cities3.xlsx")
```
Another job in Excel is to remove the sheets. For example, to remove the third sheet, use `removeSheet()` with the workbook and the sheet name or sheet number as arguments:
```{r}
removeSheet(book, sheet = "Y2010")
saveWorkbook(book, file = "cities4.xlsx")
```
These are pretty basic operations that can be easily done in Excel, but the cool thing is you can program all of these tasks in R in a reproducible way.

Build a connection to the *urbanpop.xlsx* file in R:
```{R}
my_book <- loadWorkbook("urbanpop.xlsx")
```
Check again the class of `my_book`:
```{r}
class(my_book)
```
List the sheets in `my_book`:
```{r}
sheets <- getSheets(my_book)
```
Have an overview about *urbanpop.xlsx* without having to open up the Excel file:
```{r}
all <- lapply(sheets, readWorksheet, object = my_book)
str(all)
```
Suppose we're only interested in urban population data of the years 1968, 1969 and 1970. The data for these years is in the columns 3, 4, and 5 of the second sheet. 
```{r}
urbanpop_sel <- readWorksheet(my_book, sheet = 2, 
                              startCol = 3, 
                              endCol =5)
```
`urbanpop_sel` no longer contains information about the countries now. Import first column from second sheet in `my_book`:
```{r}
countries <- readWorksheet(my_book, 
                           sheet = 2, startCol = 1, endCol = 1)
```
Joining `urbanpop_sel` and `countries` together using `cbind()`:
```{r}
selection <- cbind(countries, urbanpop_sel)
```
Where **readxl** and **gdata** are only able to import Excel data, **XLConnect**'s approach of providing an actual interface to an Excel file makes it able to edit your Excel files from inside R. 
```{r}
createSheet(my_book, name = "data_summary")
getSheets(my_book)
```
Create a data frame with some summary statistics on the two Excel sheets:
```{r}
sheets <- getSheets(my_book)[1:3]
dims <- sapply(sheets, function(x) dim(readWorksheet(my_book, sheet = x)), USE.NAMES = FALSE)
summ <- data.frame(sheets = sheets,
                   nrows = dims[1, ],
                   ncols = dims[2, ])
```
Use `writeWorksheet()` to populate the "data_summary" sheet with the `summ` data frame:
```{r}
writeWorksheet(my_book, summ, sheet = "data_summary")
```
Call `saveWorkbook()` to store the adapted Excel workbook as a new file:
```{r}
saveWorkbook(my_book, file = "summary.xlsx")
```
"data_summary" is not an ideal name. As the summary of these Excel sheets is always data-related, you simply want to name the sheet "summary".
```{r}
renameSheet(my_book, "data_summary", "summary")
```
Print out sheets of `my_book`:
```{r}
getSheets(my_book)
```
Finally, make sure to actually save the `my_book` object to a new Excel file:
```{r}
saveWorkbook(my_book, file = "renamed.xlsx")
```
Remove the fourth sheet:
```{r}
removeSheet(my_book, sheet = 4)
```
Save the resulting workbook to a file "clean.xlsx":
```{r}
saveWorkbook(my_book , file = "clean.xlsx")
```
