---
title: "IMPORTING EXCEL DATA"
source: DataCamp
---

### 3.1 Relational databases
Many companies store their information in relational databases. Take this database called "company" for example. It contains three tables: employees, products and sales. Like a flat file, information is displayed in a table format. 

-    The *employees* table has 5 records and 3 fields: *id*, *name*, and *start_at*. The *id* here serves as a unique key for each record.
-    The *products* table contains the details on 4 products. Also here each product has an identifier.  
-    The *sales* table lists which product was sold by whom, when, and for what price. The *id* in *employee_id* corresponds to the ids in the *employees* and *products* tables respectively.
    
<img src = "https://raw.githubusercontent.com/LAnalyt/R-00_Base-R/master/2_resources/import_relational%20data.JPG" width = "450">

This relation makes the database very powerful. The information is stored in separate tables whose identifier can be connected to each other through a database management system (DBMS). 

-    Open source DBMS: **MySQL**, **postgreSQL**, **SQLite**...
-    Proprietary DBMS: **Oracle Database** and **Microsoft SQL Server**

Practically all DBMS use *SQL* language for querying and maintaining database. Depending on the type of database you want to connect to, you will have to use different packages.

-    **RMySQL** package: for **MySQL** database
-    **RPostgresSQL** package: for **PostgresSQL** database
-    **ROracle** package: for **Oracale** database.

But before implementing these packages, we will need to install and load the **DBI** package for the interface. **DBI** specifies which R functions to access and manipulate the database.
```{r}
library(DBI)
```

### 3.2 Connecting to a database
The first step is to create a connection to the remote **MySQL** database.
```{r}
con <- dbConnect(RMySQL::MySQL(),    # construct  SQL driver
                 dbname = "tweater", # specify database's name
                 host =  "courses.csrrinzqubik.us-east-1.rds.amazonaws.com",
                 port = 3306,
                 user = "student",   # credentials to identify    
                 password = "datacamp") 
```
The first argument specifies the driver that you will use to connect to the **MySQL** database. The `MySQL()` from  **RMySQL** simply constructs a driver that `dbConnect` can use. Next, you have to specify the database name, where it is hosted, through which port you want to connect, and finally the credentials to authenticate yourself.
The result of `dbConnect()` is a *DBIConnection* object. This object has to be passed to any function for interacting with the database.
```{r}
class(con)
```

### 3.3 Import table data
After connecting successfully to a database, we want to see what is in there. The first step is listing all the tables in the database.
```{r}
tables <- dbListTables(con)
tables
```
We get a character vector of length 3, corresponding to the table names in the database. The database contains data on a more tasty version of Twitter, namely `Tweater`. Users can post "tweats" with short recipes for delicious snacks. People can comment on these "tweats". 
```{r}
str(tables)
```
Next, you can choose to actually read the data from one of these tables with `dbReadTable()`, for example data on the users:
```{r}
dbReadTable(con, "users") # specify table in second argument.
```
The result is a data frame, with exactly the same content as the original database table. 
Use `lapply()` to import all the tables:
```{r}
lapply(tables, dbReadTable, conn = con)
```
The "tweats" table contains a column `user_id`. The ids in the column refer to the users that have posted the tweat. Similarly, the comments contain both a `user_id` and a `tweat_id` column. It specifies which user posted a comment on which tweat.

### 3.4 SQL queries from inside R
`dbReadTable()` imports the entire table, which is not optimal with large database. Selective importing only imports the elements that you actually need inside R via SQL language.
SQL Querries allows you to retrieve data based on specific criteria. You can send these querries through R functions specified by the **DBI** package and implemented by RMySQL in this case.

<img src = "https://raw.githubusercontent.com/LAnalyt/R-00_Base-R/master/2_resources/import_dbConnect.JPG" width = "300">

```{r}
dbGetQuery(con, "SELECT name FROM users WHERE id = 1")
```
The syntax is common in SQL query, with keywords: `SELECT`, `FROM` and `WHERE`. 

-    `SELECT`: specifies which column to select
-    `FROM`: specifies which table you want to get data
-    `WHERE`: specifies a condition that a record in the table has to meet. Notice in SQL language `=` replaces `==`  conditional constraint.
Without `dbGetQuerry()`, we will have to load the entire dataset into R then create a subset out of it.
```{r}
users <- dbReadTable(con, "users")
subset(users, subset = id == 1,
       select = name)
```
The result is exactly the same.

Apart from checking equality, you can also check for less than and greater than relationships, with `<` and `>`, just like in R.
```{R}
dbGetQuery(con, "SELECT post FROM tweats WHERE date > 2015-09-21")
```
Add more conditions with `AND`:
```{r}
dbGetQuery(con, "SELECT message FROM comments WHERE tweat_id = 77 AND user_id > 4")
```
There are also dedicated SQL functions that you can use in the WHERE clause of an SQL query. For example, `CHAR_LENGTH()` returns the number of characters in a string.
```{r}
dbGetQuery(con, "SELECT id, name FROM users WHERE CHAR_LENGTH(name) < 5")
```
Another very often used keyword is `JOIN`, and more specifically `INNER JOIN`.
```{r}
dbGetQuery(con, "SELECT post FROM tweats INNER JOIN comments on tweats.id = tweat_id WHERE tweat_id = 77")
```
Under the hood of `dbGetQuery()`, R sends a query to the database with `dbSendQuery()`.
```{r}
names <- dbSendQuery(con, "SELECT name FROM users")
```
This function returns a result, but does not contain any records. For that we need to use `dbFetch()`.
```{r}
dbFetch(names)
```
Clear the result manually:
```{r}
dbClearResult(names)
```
The combination of `dbSendQuery`, `dbFetch` and `dbClearResult` gives exactly the same result as `dbGetQuery` does. But `dbFetch` query calls allow you to specify a maximum number of records to retrieve per fetch.
```{r}
names <- dbSendQuery(con, "SELECT name FROM users")
dbFetch(names, n = 2)
```
This is useful when you need to load tons of records, but want to do this chunk by chunk.
Suppose you want to want to get the result of the query record by record, you can also construct a `while` loop.
```{r}
while(!dbHasCompleted(names)) {
  chunk <- dbFetch(names, n = 1)
  print(chunk)
}
```
Create the data frame `long_tweats` where the character length of the post variable exceeds 40:
```{r}
long_tweats <- dbGetQuery(con, "SELECT post, date FROM tweats WHERE length(post) > 40")
long_tweats
```
Finally, explicitly disconnect the database after you are done:
```{R}
dbDisconnect(con)
```
